Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-10-06 paper/tidy-missing-data-paper.Rmd
JSS review

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
bibliography: "tidy-missing-data-paper.bib"
cls: jss.cls
---

@@

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} [@tidyverse]. These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains. For example, the \pkg{tidytext} [@tidytext] package for text analysis, the \pkg{tsibble} [@wang2020tsibble] package for time series data, and \pkg{tidycensus} [@tidycensus] for working with US census and boundary data.

@@

There are many ways each value can be missing, we adopt the notation used in [@VanBuuren2012]. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}". So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns. Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

The \pkg{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category. \pkg{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The \pkg{MANET} software is no longer maintained and cannot be installed. The approach of [@Swayne1998] in the software \pkg{XGobi}, further developed in \pkg{ggobi} [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in \pkg{ggobi} by setting them to be 10% below the minimum value.

@@

A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used. A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively. This matrix was used to explore missing values in the interactive graphics library \pkg{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness. The shadow matrix could be interactively linked to the data. However, there are some limitations to the shadow matrix. Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.

@@

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (\@ref(verbs-recode)).

@@

Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed. However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if} as in \@ref(verbs-replace-with). \code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with *nabular* data covered in Section \@ref(verbs-track).

@@

Missing values are often ignored when plotting data - which is why the data visualization software, \pkg{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the *nabular* data structure.

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "\\texttt{miss\\char`_var\\char`_summary} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The track missings pattern is applied (described in \@ref(verbs-track)), to assess imputed values. Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean.

@@

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]). Evaluating imputed values using a testing framework like [@VanBuuren2012] is also supported.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} [@tidyverse]. These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains. For example, the \pkg{tidytext} [@tidytext] package for text analysis, the \pkg{tsibble} [@wang2020tsibble] package for time series data, and \pkg{tidycensus} [@tidycensus] for working with US census and boundary data.

@@

There are many ways each value can be missing, we adopt the notation used in [@vanBuuren2018]. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}". So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in Section \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns. Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@vanBuuren2018]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

The \proglang{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category. \proglang{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The \proglang{MANET} software is no longer maintained and cannot be installed. The approach of [@Swayne1998] in the software \proglang{XGobi}, further developed in \proglang{ggobi} [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in \proglang{ggobi} by setting them to be 10% below the minimum value.

@@

A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used. A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively. This matrix was used to explore missing values in the interactive graphics library \proglang{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness. The shadow matrix could be interactively linked to the data. However, there are some limitations to the shadow matrix. Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.

@@

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).

@@

Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed. However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if}. \code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with *nabular* data covered in the following section.

@@

Missing values are often ignored when plotting data - which is why the data visualization software, \proglang{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the *nabular* data structure.

@@

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "\\texttt{miss\\char`_var\\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
    digits = 1)
    # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The track missings pattern is applied (described in \@ref(verbs)), to assess imputed values. Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean.

@@

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]). Evaluating imputed values using a testing framework like [@vanBuuren2018] is also supported.
